### 2017-05-23 01-intro.tex
Work on Henri's comments for 1. Introduction

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent
work~\cite{LiAW15} showed that this methodology is faster and more accurate than the
SVB method. \cite{LiAW15} further proposes a heuristic to scale up the number of
communities at the cost of less precision; the work in this paper considers the
algorithm without that heuristic.

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms, like deep learning, in several
ways. First, it is highly data-intensive which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
community graphs.

Starting from an existing implementation in Python/Numpy, we will present
how this algorithm can be implemented much more efficiently in C++. Then,
we will describe our two key contributions: first, how to scale up our
implementation to run on GPUs and multicore CPUs, second, to scale out
our implementation to overcome the memory bounds posed by GPUs or
single-machine systems in general.
% To this end, we propose a design of a parallel and distributed system specifically tailored to solve the a-MMSB problem. In particular, we use a mixture of OpenMP, MPI and RDMA in order to efficiently scale and accelerate the algorithm's computation.

The focus of the first part of our work is on the computational efficiency and parallel
performance of the SG-MCMC algorithm by
developing aggressive optimizations targeting
multi-core CPUs and GPUs. The GPU implementation achieves speedup factors up to~86,
compared to the
well-tuned sequential C++ program which itself is a factor~1000-1500 faster than
the original Python/Numpy program.
%
Through careful analysis of the computation and data structures we show that
the algorithm's full state can be reduced by roughly 75\%. Compressing the
state significantly reduces the data intensity and allows for tackling larger
problems. % while maintaining all state in memory.

@@

The second part of our work, scaling out to the largest publicly available
community graphs, also necessitated overcoming several challenges.
The algorithm's state grows rapidly with larger graphs and number of latent
communities.  Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned and distributed across a cluster
of machines. Hence, to access the full state, each cluster node must read
remote memory hosted by its peers. We leveraged RDMA to improve the latency and
bandwidth
of such operations. To reduce the
latency further, we pipelined the algorithm's computations by fetching
data in advance over the network.
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

The remainder of this paper is organized as follows.
Section~\ref{sec-background} provides an overview of the
algorithm and its theoretical foundation, with related work in
Section~\ref{sec-related}. Section~\ref{sec-gpu} delves
into the parallelization optimizations for GPUs and multicore CPUs.
Section~\ref{sec-distr} describes the design, implementation and evaluation of
the scalable distributed solution. Finally, Section~\ref{sec-conclusion}
provides concluding remarks.

>>>>

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent
work of the authors of this paper~\cite{LiAW15} showed that this methodology is faster and more accurate than the
SVB method. That work further proposes a heuristic to scale up the number of
communities at the cost of less precision; the work in this paper considers the
algorithm without that heuristic.

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms like deep learning in several
ways. First, it is \emph{highly data-intensive} which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
community graphs.

Starting from an existing implementation in Python/Numpy, we create an efficient
baseline version in C++ to serve as the baseline for the parallel versions.
This already improves performance by a factor of 1000 to 1500 compared to
Python. Then, our key contributions are:
\begin{itemize}
\item
our implementation is scaled up to run on GPUs.
We developed aggressive optimizations to solve the memory bandwidth issues for
GPUs. This implementation achieves speedup factors up to~86,
compared to the baseline.
%
Through careful analysis of the computation and data structures we show that
the algorithm's full state can be reduced by roughly 75\%. Compressing the
state significantly reduces the data intensity and allows for tackling larger
problems; % while maintaining all state in memory.

\item
our implementation is scaled out to overcome the memory bounds posed by GPUs or
single-machine systems in general, to solve the largest publicly available
community graphs. % This necessitated overcoming several challenges.
% The algorithm's state grows rapidly with larger graphs and number of latent
% communities.
Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned across a cluster
of machines. Hence, the cluster nodes must read
remote memory hosted by their peers. Since the algorithm is memory-bound, we
substantially optimized remote memory accesses by implementing
a custom RDMA store.
On top of that, we pipelined the algorithm stages by fetching
data in advance over the network;
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

\item we argue that it is valid to use a CPU version with straightforward
intra-node thread parallelism for the distribibuted implementation, and
that it is pointless to combine the GPU work into it.

\end{itemize}
% To this end, we propose a design of a parallel and distributed system specifically tailored to solve the a-MMSB problem. In particular, we use a mixture of OpenMP, MPI and RDMA in order to efficiently scale and accelerate the algorithm's computation.

@@

A bird's eye-view of the work in this paper, without any technical detail, has been
published at CCGrid-2016~\cite{10.1109/CCGrid.2016.98}.
The results of our distributed MCMC-aMMSB implementation
have been previously published at the 2016 ParLearning
workshop~\cite{DBLP:conf/ipps/El-HelwHLAWB16}, where it earned a best-paper
award.

The remainder of this paper is organized as follows.
Section~\ref{sec-background} provides an overview of the
algorithm and its theoretical foundation, with related work in
Section~\ref{sec-related}. Section~\ref{sec-design-sequential} describes how we
arrived at an efficient sequential version that serves as the baseline for the
parallel implementations. Section~\ref{sec-gpu} delves
into the parallelization optimizations for GPUs and multicore CPUs.
Section~\ref{sec-distr} describes the design, implementation and evaluation of
the scalable distributed solution. Finally, Section~\ref{sec-conclusion}
provides concluding remarks.
